Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-08 paper-amia-summit-2018/amia_summit_2018.tex
update most likely sequences and T-SNE diagram

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 4225 & 82.20\% & 10.05 \\\hline
Unsuccessful sequences & 918 & 17.80\% & 9.89 \\\hline
  \end{tabular}
\end{table}

@@

@@

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds. However, LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). However, Accuracy is simply computed as the ratio of correctly classified sequences and a total number of sequences.

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences. \\

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7013 & 0.7037 & 0.7013 & 0.7003\\ \hline
 Markov Chain $2^{nd}$ Order & 0.5905 & 0.5917 & 0.5905 & 0.5890\\ \hline
 Hidden Markov Model & 0.5439 & 0.5778 & 0.5439 & 0.4865\\ \hline
 LSTM RNN & 0.8770 & 0.8812 & 0.8770 & 0.8766\\ \hline
 LSTM RNN - TR & 0.8824 & 0.8890 & 0.8824 & 0.8817\\ \hline
 GRU RNN & 0.8844 & 0.8897 & 0.8844 & 0.8840\\ \hline
 GRU RNN - TR & \textbf{0.8858} & \textbf{0.8917} & \textbf{0.8858} & \textbf{0.8853}\\ \hline 
  \end{tabular}
\end{table}

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.00005 and early stopping strategy with 150 epoch for training deep learning models when under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, recurrent neural networks outperform over probabilistic models and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower accuracy, precision, recall, and F-measure than first-order MC. In particular, accuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models. \\

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7930 & 0.8127 & 0.7930 & 0.7896\\ \hline
 Markov Chain $2^{nd}$ Order & 0.7325 & 0.7461 & 0.7325 & 0.7288\\ \hline
 Hidden Markov Model & 0.7763 & 0.8062 & 0.7763 & 0.7706\\ \hline
 LSTM RNN & 0.8703 & 0.8782 & 0.8703 & 0.8696\\ \hline
 LSTM RNN - TR & 0.8713 & 0.8789 & 0.8713 & 0.8706\\ \hline
 GRU RNN & 0.8711 & 0.8785 & 0.8711 & 0.8705\\ \hline
 GRU RNN - TR & \textbf{0.8722} & \textbf{0.8790} & \textbf{0.8722} & \textbf{0.8716}\\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy. Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences. These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still has 2.41\% lower F1-score than first-order Markov chain. Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences whereas the performance of probabilistic models decreases. We also observed that all models have the largest value in precision compared to other performance metrics.

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation. Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews. \\

\begin{table}[h]
\centering
\caption{Common patterns in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Pattern} \\ \hline      
successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
  \end{tabular}
\end{table}

@@

\section*{Acknowledgments}
This study was supported by an R21 grant DK108071 from the NIH. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help with transcribing the recordings of motivational interviews. 
\pagebreak

>>>>

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 4225 & 82.15\% & 9.79 \\\hline
Unsuccessful sequences & 918 & 17.85\% & 9.65 \\\hline
  \end{tabular}
\end{table}

@@

\textbf {MYSCOPE Code Embedding}: is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. Stochastic Neighbor Embedding for MYSCOPE code is shown in Figure~\ref{fig:code_embedding}.
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}
    \label{fig:code_embedding}
\end{figure}

@@

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds. However, LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)).

\section*{Results}
Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.
\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under- and over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under Sampling}} & \multicolumn{3}{|c|}{\textbf{Over Sampling}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7060 & 0.7044 & 0.7038 & 0.7932 & 0.7799 & 0.7775 \\ \hline
 Markov Chain $2^{nd}$ Order & 0.6395 & 0.6385 & 0.6379 & 0.7111 & 0.7029 & 0.7000\\ \hline
 Hidden Markov Model & 0.6244 & 0.6143 & 0.6067 & 0.7775 & 0.7567 & 0.7520\\ \hline
 LSTM RNN & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
 LSTM RNN - TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
 GRU RNN & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
 GRU RNN - TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
  \end{tabular}
\end{table}

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.00005 and early stopping strategy with 150 epoch for training deep learning models when under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform over probabilistic models and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower accuracy, precision, recall, and F-measure than first-order MC. In particular, accuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models.

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy. Table~\ref{tab:result_under_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences. These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still has 2.41\% lower F1-score than first-order Markov chain. Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences whereas the performance of probabilistic models decreases. We also observed that all models have the largest value in precision compared to other performance metrics.

\subsection*{\textit{Most likely communication sequences}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful sequence start with a summary of the discussion or structured statement. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, providing additional information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation. Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews.
\begin{table}[h]
\centering
\caption{Most likely communication sequences in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Most likely communication sequences} \\ \hline      
successful & GINFO+: General information, positive $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ OQTBN: \\ 
& Open-ended question, target behavior neutral \\\hline
successful & SS: Structure session $\rightarrow $ GINFO+: General information, positive $\rightarrow $ CQECHT+: Closed-ended \\
& question, elicit change talk positive\\\hline
successful & SO: Statement, other $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ AF: Affirm $\rightarrow $ HUPW: High uptake, \\
& weight $\rightarrow $ OQECML+: Open-ended question, elicit commitment language positive. \\\hline
unsuccessful & ADV+: Advise, positive $\rightarrow $ AMB-: Ambivalence negative $\rightarrow $ OQECHT-: Open-ended \\
& question, elicit change talk negative\\\hline
unsuccessful & CQECHT+: Open-ended question, elicit change talk positive $\rightarrow $ RCHT-S: Reflect, change talk \\
& negative $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative \\\hline
unsuccessful & SUP: Support $\rightarrow $ AF: Affirm $\rightarrow $ CQTBN: Closed-ended question, target behavior neutral \\
& $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative $\rightarrow $ AMB-: Ambivalence negative\\\hline
  \end{tabular}
\end{table}

@@

\section*{Acknowledgments}
This study was supported by an R21 grant DK108071 from the NIH. We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help with transcribing the recordings of motivational interviews. 
%\pagebreak
